A Classical Density-Functional Theory for Describing Water Interfaces

### 2013-08-08 papers/water-saft/paper.tex
Small changes to water paper Tested everything on node25.

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)\cite{jeanmairet2013molecular,
  zhao2011molecular, zhao2011new, ramirez2005direct,
  ramirez2005density, levesque2012solvation, levesque2012scalar}
~\fixme{These are from Jeanmairet 2013 citations 35-44; only the ones
  that discuss water}.  Classical DFT is based on a description of a fluid written
as a free energy functional of the density distribution.  There are
two approaches used to construct a classical DFT for water.  The first
is to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure
\cite{jeanmairet2013molecular} ~\fixme{cite Jeanmairet and similar}.
Using this approach, it is possible to construct a functional that
reproduces the exact second-order response function of the liquid
under the fitted conditions.  This class of functional, however, will
be less accurate at other temperatures or pressures---and in the
inhomogeneous scenarios in which solvation models are applied.  The
second approach is construct a functional by applying liquid-state
theory to a model system, and then fit the model to experimental data
such as the equation of state.

@@

We modify our previous association free energy term to use our
$g_{\sigma}^\textit{A}$ rather than Yu and Wu's $g_\sigma$.  The
association free energy for our four-site model has the form:
\begin{align}
  F_\text{assoc}[n] &= k_BT \int n_\text{site}(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where
\begin{align}
  n_\text{site}(\rr) &=
  \begin{cases}
    4 n(\rr) & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites at position~$\rr$.  The four in
$n_\text{site}(\rr)$ comes from the four hydrogen bond sites,
$n_0(\rr)$ is the average density contacting point $\rr$, and
$\zeta(\xx)$ is a dimensionless measure of the density inhomogeneity
from Yu and Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}.  The
functional $X(\rr)$ is the fraction of association sites \emph{not}
hydrogen-bonded, which is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 2n_\text{site}'(\rr)
      \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)} - 1}
  {n_\text{site}'(\rr)
    \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)}, \label{eq:X}
\end{align}
where
\begin{align}
  n_\text{site}'(\rr) &=
  \begin{cases}
    \frac{4}{\pi\sigma^2} \int n(\rr')\delta(\sigma - |\rr-\rr'|) d\rr' & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites that could bond to the sites~$n_\text{site}(\rr)$, and
\begin{align}
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The radial distribution function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$. Four our old functional
we used Yu and Wu's radial distribution function, and for this paper we
use our $g^\textit{A}_\sigma$.

@@

We first simulate a single hydrophobic rod immersed in water. The
density profile of the water with our new functional is compared
agains our previous results in Fig. \ref{fig:density-single-rod}. The
density at contact with the rods are similar and there are slight
difference in density, particularly near the first peak after contact;
all rods we tested exhibited this sharp peak near one diamater form
contact. Since we have changed the association term in the functional,
a difference will occur in the average number of hydrogen bonds. We
compare the hydrogen bonds per length with the new and old functionals
for various rods (shown in Fig. \ref{fig:single-rod-broken-HB}) and
find that there is a noteable difference. The number of broken bonds
both increase linearly with the rod radius, but where this starts
differs between the two methods.

@@

We continue our study in three dimensions with a hard sphere solute in
water. Similar to the single rod, we see a decrease in the number of
broken hydrogen bonds for a given rod radius. However, with the single
rods the number of hydrogen bonds appears to increase linearly with
rod radius where the sphere radii we used. \fixme{When the size of
  each rod is increased, the surface area increases proportional to
  radius.  However, when we increase the radius of each sphere, the
  surface area increases as $R^2$.}

>>>>

Water, the universal solvent, is of critical practical importance, and
a continuum description of water is in high demand for a solvation
model.  A number of recent attempts to develop improved solvation
models for water have built on the approach of classical
density-functional theory (DFT)~\cite{jeanmairet2013molecular,
  zhao2011molecular, zhao2011new, ramirez2005direct,
  ramirez2005density, levesque2012solvation, levesque2012scalar}.
Classical DFT is based on a description of a fluid written as a free
energy functional of the density distribution.  There are two
approaches used to construct a classical DFT for water.  The first is
to choose a functional form that is fit to properties of the bulk
liquid at a given temperature and pressure
\cite{jeanmairet2013molecular} ~\fixme{cite Jeanmairet and similar}.
Using this approach, it is possible to construct a functional that
reproduces the exact second-order response function of the liquid
under the fitted conditions.  This class of functional, however, will
be less accurate at other temperatures or pressures---and in the
inhomogeneous scenarios in which solvation models are applied.  The
second approach is construct a functional by applying liquid-state
theory to a model system, and then fit the model to experimental data
such as the equation of state.

@@

We modify our previous association free energy term to use our
$g_{\sigma}^\textit{A}$ rather than Yu and Wu's $g_\sigma$.  The
association free energy for our four-site model has the form:
\begin{align}
  F_\text{assoc}[n] &= k_BT \int n_\text{site}(\xx)
  \left(\ln X(\xx) - \frac{X(\xx)}{2} + \frac12\right) d\xx
\end{align}
where
\begin{align}
  n_\text{site}(\rr) &=
  \begin{cases}
    4 n(\rr) & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites at position~$\rr$.  The four in
$n_\text{site}(\rr)$ comes from the four hydrogen bond sites,
$n_0(\rr)$ is the average density contacting point $\rr$, and
$\zeta(\xx)$ is a dimensionless measure of the density inhomogeneity
from Yu and Wu~\cite{yu2002fmt-dft-inhomogeneous-associating}.  The
functional $X(\rr)$ is the fraction of association sites \emph{not}
hydrogen-bonded, which is determined by the quadratic equation
\begin{align}
  X(\xx) &= \frac{\sqrt{1 + 2n_\text{site}'(\rr)
      \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)} - 1}
  {n_\text{site}'(\rr)
    \kappaassoc g^\textit{SW}_\sigma(\xx)
  \left(e^{-\beta\epsilonassoc} - 1\right)}, \label{eq:X}
\end{align}
where
\begin{align}
  n_\text{site}'(\rr) &=
  \begin{cases}
    \frac{4}{\pi\sigma^2} \int n(\rr')\delta(\sigma - |\rr-\rr'|) d\rr' & \text{this work}\\
    4 n_0(\rr) \zeta(\rr) & \text{Ref.~\citenum{hughes2013classical}}
  \end{cases} 
\end{align}
is the density of bonding sites that could bond to the sites~$n_\text{site}(\rr)$, and
\begin{align}
  g^\textit{SW}_\sigma(\xx) &= g^\textit{HS}_\sigma(\xx) +
  \frac{1}{4}\beta\left(\frac{\partial a_1}{\partial \etadisp(\xx)} -
  \frac{\lambdadisp}{3 \etadisp}\frac{\partial a_1}{\partial \lambdadisp}\right)\label{eq:gSW},
\end{align}
%This is equation 77 of gil-villegas-1997-SAFT-VR
where $g^\textit{SW}_\sigma$ is the correlation function evaluated at
contact for a hard-sphere fluid with a square-well dispersion
potential, and $a_1$ and $a_2$ are the two terms in the dispersion
free energy.  The radial distribution function $g^\textit{SW}_\sigma$ is
written as a perturbative correction to the hard-sphere fluid
correlation function $g^\textit{HS}_\sigma$. Four our old functional
we used Yu and Wu's radial distribution function, and for this paper we
use our $g^\textit{A}_\sigma$.

@@

We first simulate a single hydrophobic rod immersed in water. The
density profile of the water with our new functional is compared
agains our previous results in Fig. \ref{fig:density-single-rod}. The
density at contact with the rods are similar and there are slight
difference in density, particularly near the first peak after contact;
all rods we tested exhibited this sharp peak near one diamater from
contact. Since we have changed the association term in the functional,
a difference will occur in the average number of hydrogen bonds. We
compare the hydrogen bonds per length with the new and old functionals
for various rods (shown in Fig. \ref{fig:single-rod-broken-HB}) and
find that there is a noteable difference. The number of broken bonds
both increase linearly with the rod radius, but where this starts
differs between the two methods.

@@

A common test case for studying hydrophobic solutes in water is the
hard-sphere solute. We continue our study in three dimensions by
creating a hydrophobic spherical cavity in water. Similar to the
single rods, we see a decrease in the number of broken hydrogen bonds
for spheres of the same radius compared to our previous
functional. The number of broken hydrogen bonds increases linearly
with rod radius for large enough rods, where for spheres we don't see
this linearity (Fig. \ref{fig:spheres-broken-HB}). \fixme{When the
  size of each rod is increased, the surface area increases
  proportional to radius.  However, when sphere radius increase, the
  surface area increases as $R^2$.}
